---
title: "STAT 566 Final Project"
author: 'Andrew (Andy) Shin (ID: 1937870)'
date: "2026-04-27"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(haven)
library(dplyr)

library(dagitty)
library(ggdag)
library(ggplot2)

```


```{r}
rhc <- read.csv("../data/processed/rhc-processed-data.csv")

# rhc$death30d <- factor(rhc2$death30d)
# rhc$RHC <- factor(rhc$RHC)
```


```{r}
# Install necessary packages if you haven't already
# install.packages(c("dagitty", "ggdag", "ggplot2"))
# Define the DAG and set coordinates to match the image layout
rhc_dag <- dagitty("dag {
  cat1 -> pafi1;
  cat1 -> scoma1;
  cat1 -> RHC;
  age -> chfhx;
  age -> dnr1_Yes;
  age -> meanbp1;
  chfhx -> death30d;
  chfhx -> meanbp1;
  chfhx -> pafi1;
  meanbp1 -> aps1;
  meanbp1 -> death30d;
  meanbp1 -> pafi1;
  meanbp1 -> scoma1;
  meanbp1 -> RHC;
  pafi1 -> aps1;
  pafi1 -> dnr1_Yes;
  pafi1 -> RHC;
  scoma1 -> aps1;
  scoma1 -> dnr1_Yes;
  aps1 -> death30d;
  aps1 -> RHC;
  dnr1_Yes -> death30d;
  RHC -> death30d
}")

coordinates(rhc_dag) <- list(
  x = c(cat1 = 0, age = 0, chfhx = 0, 
        pafi1 = 1, meanbp1 = 1, scoma1 = 1, 
        aps1 = 2, dnr1_Yes = 2, 
        RHC = 3, death30d = 3),
  
  # Staggered Y-coordinates to dodge straight lines
  y = c(cat1 = 0, age = 2, chfhx = 4, 
        pafi1 = -1, meanbp1 = 2, scoma1 = 5, 
        aps1 = 0.5, dnr1_Yes = 3.5, 
        RHC = 0, death30d = 2.5)
)

# 1. Base dagitty plot
plot(rhc_dag)

# 2. Presentation-ready plot using ggdag
ggdag(rhc_dag, text = FALSE, use_labels = "name") +
  theme_dag() +
  ggtitle("Figure 2: DAG model of RHC study") +
  theme(plot.title = element_text(hjust = 0.5))
```

RHC: Right Heart Catheterization (RHC) status during the first 24 hours (Treatment, $A$).

death30d: All-cause mortality up to 180 days (Outcome, $Y$).

age: Patient age in years.

cat1: Primary disease category prompting ICU admission.

chfhx: Baseline history of Congestive Heart Failure. 

meanbp1: Mean blood pressure recorded on day 1.

pafi1: Ratio of arterial oxygen partial pressure to fractional inspired oxygen ($\text{PaO}_2/\text{FIO}_2$) on day 1.

scoma1: Glasgow Coma Score on day 1.

aps1: APACHE III score on day 1.

dnr1_yes: Do Not Resuscitate (DNR) status established on day 1.


```{r}
adjustmentSets(rhc_dag, exposure = "RHC", outcome = "death30d")
```


```{r}
library(ggplot2)

# Ensure death30d is treated as a categorical factor, not a continuous integer
rhc$death30d <- factor(rhc$death30d, levels = c(0, 1))

ggplot(rhc, aes(x = scoma1, fill = factor(RHC))) +
  # alpha = 0.5 makes the distributions semi-transparent so you can see the overlap
  geom_density(alpha = 0.5) +
  
  # Optional: Define clean, colorblind-friendly colors and update legend labels
  scale_fill_manual(
    values = c("0" = "#E69F00", "1" = "#56B4E9"),
    labels = c("0 (Control/Died)", "1 (Treated/Survived)")
  ) +
  
  # Clean, publication-ready theme
  theme_minimal() +
  labs(
    title = "Density Distribution by death30d Status",
    x = "Glasgow Coma Score on day 1",
    y = "Density",
    fill = "death30d Status"
  ) +
  theme(legend.position = "top")
```


```{r}
rhc <- rhc %>%
  mutate(
    cat1 = case_when(
      cat1_CHF == 1             ~ "CHF",
      cat1_Cirrhosis == 1       ~ "Cirrhosis",
      cat1_Colon_Cancer == 1    ~ "Colon Cancer",
      cat1_Coma == 1            ~ "Coma",
      cat1_COPD == 1            ~ "COPD",
      cat1_Lung_Cancer == 1     ~ "Lung Cancer",
      cat1_MOSF_Malignancy == 1 ~ "MOSF Malignancy",
      cat1_MOSF_Sepsis == 1     ~ "MOSF Sepsis",
      TRUE                      ~ "ARF"  # If all above are 0, it defaults to ARF
    )
  )
```


```{r}
cat_vars <- c("aps1", "chfhx", "dnr1_Yes", "meanbp1")
death30d_form <- as.formula(paste("death30d ~ ", 
                                  paste(cat_vars, collapse = "+")))

fit_mu0 <- glm(death30d_form, data = subset(rhc, RHC == 0), family=binomial)
mu0_hat <- predict(fit_mu0, newdata = rhc, type="response")
tau_ATT_reg <- mean(rhc$death30d[rhc$RHC==1] - mu0_hat[rhc$RHC == 1])

cat("ATT via outcome regression: ", round(tau_ATT_reg, 3), "\n")
```


```{r}
e_hat <- predict(glm(RHC ~ chfhx + aps1 + meanbp1 + dnr1_Yes, data = rhc, family = binomial), type = "response")

w_treat <- rhc$RHC + (1 - rhc$RHC)  * (e_hat / (1 - e_hat))
tau_ATT_IPW <- mean(rhc$death30d[rhc$RHC == 1]) - (sum((1 - rhc$RHC) * w_treat * rhc$death30d) / sum((1-rhc$RHC) * w_treat))

cat("ATT via IPW: ", round(tau_ATT_IPW, 3), "\n")
```

```{r}
if (!requireNamespace("cobalt", quietly = TRUE)) install.packages("cobalt")
library(cobalt)

bal.plot(RHC ~ chfhx + aps1 + meanbp1 + pafi1 + scoma1, data = rhc, weights = w_treat,
         distance = e_hat, method = "weighting", estimand = "ATT",
         var.name = "distance", which = "both", sample.names = c("Unweighted", "Weighted"))

```

```{r}
if (!requireNamespace("Matching", quietly=TRUE)) install.packages("Matching")
library(Matching)

m_att <- Match(Y = rhc$death30d, Tr = rhc$RHC, X = e_hat,
estimand = "ATT", M = 1, replace = TRUE)
tau_ATT_match <- m_att$est

cat("ATT via 1:1 PS matching: ", round(tau_ATT_match, 3), "\n")
```



```{r}
tau_ATT_AIPW <- mean(
  rhc$RHC * (rhc$death30d - mu0_hat) - (1 - rhc$RHC) * 
    (e_hat / (1 - e_hat)) * (rhc$death30d - mu0_hat)) / mean(rhc$RHC)

cat("ATT via AIPW (doubly robust): ", round(tau_ATT_AIPW, 3), "\n\n")
```


```{r}
one_rep <- function(covs_formula, df) {
  # standardization regression
  cat_vars <- c("aps1", "chfhx", "dnr1_Yes", "meanbp1")
  death30d_form <- as.formula(paste("death30d ~ ", covs_formula))

  fit_mu0 <- glm(death30d_form, data = subset(df, RHC == 0), family=binomial)
  mu0_hat <- predict(fit_mu0, newdata = df, type="response")
  tau_ATT_reg <- mean(df$death30d[df$RHC==1] - mu0_hat[df$RHC == 1])
  
  # Hajek IPW
  e_hat <- predict(glm(RHC ~ chfhx + aps1 + meanbp1 + dnr1_Yes, data = df, family =
                         binomial), type = "response")

  w_treat <- df$RHC + (1 - df$RHC)  * (e_hat / (1 - e_hat))
  tau_ATT_IPW <- mean(df$death30d[df$RHC == 1]) - (sum((1 - df$RHC) * w_treat *
                                                           df$death30d) / sum((1-df$RHC)
                                                                               * w_treat))
  # Matching
  m_att <- Match(Y = df$death30d, Tr = df$RHC, X = e_hat, 
                 estimand = "ATT", M = 1, replace = TRUE)
  tau_ATT_match <- m_att$est

  
  # AIPW (doubly robust)
  tau_ATT_AIPW <- mean(df$RHC * (df$death30d - mu0_hat) - (1 - df$RHC) * 
                         (e_hat / (1 - e_hat)) * (df$death30d - mu0_hat)) / mean(df$RHC)

  return(c(reg_via_std = round(tau_ATT_reg, 3),
    Hajek_IPW = round(tau_ATT_IPW, 3),
    Match = round(tau_ATT_match, 3),
    aipw = round(tau_ATT_AIPW, 3)))
}
```



```{r}
set.seed(2026)

B <- 500
covs_formula <- paste(cat_vars, collapse = "+")

boot <- replicate(B, {
  idx <- sample(nrow(rhc), replace = TRUE)
  d <- rhc[idx, ]
  
  res <- suppressWarnings(one_rep(covs_formula, d))
  c(Reg = res[[1]], 
    IPW = res[[2]],
    Match = res[[3]],
    AIPW = res[[4]])
})

# After populating: `boot` is a 4 x B matrix.
# 95% percentile CIs:
ci <- t(apply(boot, 1, quantile, probs = c(0.025, 0.975), na.rm=TRUE))

results_p1 <- data_frame(
  Estimator = c("Reg", "IPW", "Match", "AIPW"), 
  Estimate = one_rep(covs_formula, rhc),
  CI_lower = ci[,1],
  CI_upper = ci[,2]
)

results_p1
```






