---
title: "Balance Diagnosis"
date: "`r Sys.Date()`"
output:
  html_notebook:
    theme: united
    hightlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
project_root <- if (dir.exists("estimators")) normalizePath(".") else normalizePath("..")
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
library(data.table)
library(readr)
library(ggplot2)

old_wd <- getwd()
setwd(project_root)
source(file.path(project_root, "estimators/utils.R"))
setwd(old_wd)
```

## Setup

```{r load-data}
rhc <- fread("data/processed/rhc-processed-data.csv")
var_sets <- readr::read_rds("data/processed/rhc-var-sets.rds")

treat_var <- var_sets$treatment
covariates <- var_sets$adjustment_sets$basic

ps <- fit_propensity_score(
    data = rhc,
    treat_var = treat_var,
    covariates = covariates
)

A <- rhc[[treat_var]]
rhc[, ps := ps] |> invisible()
rhc[, ate_weight := A / ps + (1 - A) / (1 - ps)] |> invisible()
rhc[, att_weight := A + (1 - A) * ps / (1 - ps)] |> invisible()
```

## Balance helpers

```{r balance-helpers}
weighted_mean <- function(x, w) {
    sum(w * x, na.rm = TRUE) / sum(w[!is.na(x)])
}

weighted_var <- function(x, w) {
    m <- weighted_mean(x, w)
    sum(w * (x - m)^2, na.rm = TRUE) / sum(w[!is.na(x)])
}

weighted_smd <- function(data, var, treat_var, weight_var = NULL) {
    A <- data[[treat_var]]
    x <- data[[var]]

    if (is.null(weight_var)) {
        m1 <- mean(x[A == 1], na.rm = TRUE)
        m0 <- mean(x[A == 0], na.rm = TRUE)
        v1 <- stats::var(x[A == 1], na.rm = TRUE)
        v0 <- stats::var(x[A == 0], na.rm = TRUE)
    } else {
        w <- data[[weight_var]]
        m1 <- weighted_mean(x[A == 1], w[A == 1])
        m0 <- weighted_mean(x[A == 0], w[A == 0])
        v1 <- weighted_var(x[A == 1], w[A == 1])
        v0 <- weighted_var(x[A == 0], w[A == 0])
    }

    pooled_sd <- sqrt((v1 + v0) / 2)
    if (is.na(pooled_sd) || pooled_sd == 0) return(0)
    abs((m1 - m0) / pooled_sd)
}
```

## Standardized mean differences

```{r balance-table}
balance <- rbindlist(lapply(covariates, function(var) {
    data.table(
        covariate = var,
        raw = weighted_smd(rhc, var, treat_var),
        ate_ipw = weighted_smd(rhc, var, treat_var, "ate_weight"),
        att_ipw = weighted_smd(rhc, var, treat_var, "att_weight")
    )
}))

balance[order(-raw)]
```

```{r love-plot, fig.width=7, fig.height=5}
balance_long <- melt(
    balance,
    id.vars = "covariate",
    variable.name = "sample",
    value.name = "abs_smd"
)
balance_long[, covariate := factor(covariate, levels = balance[order(raw), covariate])]

love_plot <- ggplot(balance_long, aes(x = covariate, y = abs_smd, color = sample)) +
    geom_point(size = 2.2) +
    geom_hline(yintercept = 0.1, color = "grey45", linetype = "dashed") +
    coord_flip() +
    labs(
        x = NULL,
        y = "Absolute SMD",
        color = "Sample",
        title = "Covariate balance before and after IPW"
    ) +
    theme_minimal()

love_plot
```

## Propensity-score overlap

```{r propensity-score-overlap, fig.width=7, fig.height=4.5}
overlap_plot <- ggplot(rhc, aes(x = ps, fill = factor(get(treat_var)))) +
    geom_density(alpha = 0.45) +
    labs(
        x = "Estimated propensity score",
        y = "Density",
        fill = "RHC",
        title = "Propensity score overlap by treatment group"
    ) +
    theme_minimal()

overlap_plot
```

## Findings

- Raw balance is summarized by absolute standardized mean differences for each covariate in the basic adjustment set.
- ATE and ATT weighting generally target different pseudo-populations, so their balance diagnostics do not need to be identical.
- The propensity-score density plot checks whether treated and control patients have enough overlap for weighting to be credible.
