---
title: "Simple Sensitivity Analysis"
author: "Liang-Cheng Chen"
date: "`r Sys.Date()`"
output:
  html_notebook:
    theme: united
    hightlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
project_root <- if (dir.exists("estimators")) normalizePath(".") else normalizePath("..")
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
library(data.table)
library(readr)
library(ggplot2)

old_wd <- getwd()
setwd(project_root)
source(file.path(project_root, "estimators/ate/ipw.R"))
source(file.path(project_root, "estimators/ate/matching.R"))
source(file.path(project_root, "estimators/att/ipw.R"))
source(file.path(project_root, "estimators/att/matching.R"))
source(file.path(project_root, "estimators/utils.R"))
setwd(old_wd)
```

## Setup

```{r load-data}
rhc <- fread("data/processed/rhc-processed-data.csv")
var_sets <- readr::read_rds("data/processed/rhc-var-sets.rds")

outcome_var <- "death30d"
treat_var <- var_sets$treatment
covariates <- var_sets$adjustment_sets$basic
```

# Robustness Checks

## Trimming sensitivity

```{r trimming-sensitivity}
trim_grid <- list(
    c(0.001, 0.999),
    c(0.010, 0.990),
    c(0.020, 0.980),
    c(0.050, 0.950),
    c(0.100, 0.900)
)

effective_sample_size <- function(w) {
    sum(w)^2 / sum(w^2)
}

trim_results <- rbindlist(lapply(trim_grid, function(trim) {
    ps <- fit_propensity_score(
        data = rhc,
        treat_var = treat_var,
        covariates = covariates,
        trim = trim
    )
    A <- rhc[[treat_var]]
    ate_weight <- A / ps + (1 - A) / (1 - ps)
    att_weight <- A + (1 - A) * ps / (1 - ps)

    rbind(
        data.table(
            estimand = "ATE",
            estimator = "Hajek IPW",
            trim_lower = trim[1],
            trim_upper = trim[2],
            estimate = estimate_ate_hajek_ipw(
                rhc, outcome_var, treat_var, covariates, trim = trim
            ),
            ess = effective_sample_size(ate_weight)
        ),
        data.table(
            estimand = "ATT",
            estimator = "Hajek IPW",
            trim_lower = trim[1],
            trim_upper = trim[2],
            estimate = estimate_att_hajek_ipw(
                rhc, outcome_var, treat_var, covariates, trim = trim
            ),
            ess = effective_sample_size(att_weight)
        )
    )
}))

trim_results[, trim := sprintf("[%.3f, %.3f]", trim_lower, trim_upper)] |> invisible()
trim_results[]
```

```{r trimming-sensitivity-plot, fig.width=7, fig.height=4.5}
trim_plot <- ggplot(trim_results, aes(x = trim, y = estimate, color = estimand, group = estimand)) +
    geom_point(size = 2.4) +
    geom_line() +
    geom_hline(yintercept = 0, color = "grey45", linetype = "dashed") +
    labs(
        x = "Propensity-score trim",
        y = "Estimated risk difference",
        color = "Estimand",
        title = "IPW sensitivity to propensity-score trimming"
    ) +
    theme_minimal()

trim_plot
```

## Adjustment-set sensitivity

```{r adjustment-set-sensitivity}
adjustment_sets <- var_sets$adjustment_sets[c("minimal", "basic", "detailed")]

set_results <- rbindlist(lapply(names(adjustment_sets), function(set_name) {
    set_covariates <- adjustment_sets[[set_name]]

    data.table(
        adjustment_set = set_name,
        estimand = c("ATE", "ATE", "ATT", "ATT"),
        estimator = c("Hajek IPW", "PS matching", "Hajek IPW", "PS matching"),
        estimate = c(
            estimate_ate_hajek_ipw(rhc, outcome_var, treat_var, set_covariates),
            estimate_ate_ps_matching(rhc, outcome_var, treat_var, set_covariates),
            estimate_att_hajek_ipw(rhc, outcome_var, treat_var, set_covariates),
            estimate_att_ps_matching(rhc, outcome_var, treat_var, set_covariates)
        ),
        n_covariates = length(set_covariates)
    )
}))

set_results[, adjustment_set := factor(adjustment_set, 
                                       levels = c("minimal", "basic", "detailed"), 
                                       ordered = TRUE)] |> invisible()

set_results[]
```

```{r adjustment-set-sensitivity-plot, fig.width=7, fig.height=4.5}
set_plot <- ggplot(
    set_results,
    aes(x = adjustment_set, y = estimate, color = estimator, group = estimator)
) +
    geom_point(size = 2.4) +
    geom_line() +
    geom_hline(yintercept = 0, color = "grey45", linetype = "dashed") +
    facet_wrap(~ estimand) +
    labs(
        x = "Adjustment set",
        y = "Estimated risk difference",
        color = "Estimator",
        title = "Sensitivity to adjustment-set choice"
    ) +
    theme_minimal()

set_plot
```

# Sensitivity Analysis (WiP)

## E-values and partial R-squared

```{r evalues}
make_evalue <- function(rr) {
    rr_bound <- ifelse(rr < 1, 1 / rr, rr)
    ifelse(rr_bound == 1, 1, rr_bound + sqrt(rr_bound * (rr_bound - 1)))
}

estimate_regression_rr <- function(data, outcome_var, treat_var, covariates) {
    fit <- glm(
        make_interacted_outcome_formula(outcome_var, treat_var, covariates),
        data = data,
        family = binomial("logit")
    )

    data1 <- copy(data)
    data0 <- copy(data)
    data1[, (treat_var) := 1]
    data0[, (treat_var) := 0]

    p1 <- mean(predict(fit, newdata = data1, type = "response"))
    p0 <- mean(predict(fit, newdata = data0, type = "response"))
    p1 / p0
}

estimate_ate_ipw_rr <- function(data, outcome_var, treat_var, covariates, trim = c(0.02, 0.98)) {
    A <- data[[treat_var]]
    Y <- data[[outcome_var]]
    ps <- fit_propensity_score(data, treat_var, covariates, trim = trim)

    w1 <- A / ps
    w0 <- (1 - A) / (1 - ps)
    p1 <- sum(w1 * Y) / sum(w1)
    p0 <- sum(w0 * Y) / sum(w0)
    p1 / p0
}

estimate_att_ipw_rr <- function(data, outcome_var, treat_var, covariates, trim = c(0.02, 0.98)) {
    A <- data[[treat_var]]
    Y <- data[[outcome_var]]
    ps <- fit_propensity_score(data, treat_var, covariates, trim = trim)

    w0 <- (1 - A) * ps / (1 - ps)
    p1 <- mean(Y[A == 1])
    p0 <- sum(w0 * Y) / sum(w0)
    p1 / p0
}

A <- rhc[[treat_var]]
Y <- rhc[[outcome_var]]

evalue_results <- data.table(
    estimand = c("Raw", "ATE", "ATE", "ATT"),
    estimator = c("Naive risk ratio", "Regression adjustment", "Hajek IPW", "Hajek IPW"),
    risk_ratio = c(
        mean(Y[A == 1]) / mean(Y[A == 0]),
        estimate_regression_rr(rhc, outcome_var, treat_var, covariates),
        estimate_ate_ipw_rr(rhc, outcome_var, treat_var, covariates),
        estimate_att_ipw_rr(rhc, outcome_var, treat_var, covariates)
    )
)
evalue_results[, e_value := make_evalue(risk_ratio)] |> invisible()

evalue_results[]
```

```{r evalue-plot, fig.width=7, fig.height=4.5}
evalue_results[, label := paste(estimand, estimator, sep = ": ")] |> invisible()
evalue_results[, label := factor(label, levels = rev(label))]|> invisible()

evalue_plot <- ggplot(evalue_results, aes(x = label, y = e_value, color = estimand)) +
    geom_point(size = 2.6) +
    geom_hline(yintercept = 1, color = "grey45", linetype = "dashed") +
    coord_flip() +
    labs(
        x = NULL,
        y = "E-value",
        color = "Estimand",
        title = "E-values for risk-ratio estimates"
    ) +
    theme_minimal()

evalue_plot
```

```{r partial-r2}
make_lm_formula <- function(lhs, rhs_vars) {
    if (length(rhs_vars) == 0) {
        as.formula(paste(lhs, "~ 1"))
    } else {
        reformulate(rhs_vars, response = lhs)
    }
}

partial_r2_from_models <- function(full_model, reduced_model) {
    full_r2 <- summary(full_model)$r.squared
    reduced_r2 <- summary(reduced_model)$r.squared
    max(0, (full_r2 - reduced_r2) / (1 - reduced_r2))
}

treatment_outcome_partial_r2 <- partial_r2_from_models(
    lm(make_lm_formula(outcome_var, c(treat_var, covariates)), data = rhc),
    lm(make_lm_formula(outcome_var, covariates), data = rhc)
)

benchmark_r2 <- rbindlist(lapply(covariates, function(var) {
    other_covariates <- setdiff(covariates, var)

    treatment_full <- lm(make_lm_formula(treat_var, covariates), data = rhc)
    treatment_reduced <- lm(make_lm_formula(treat_var, other_covariates), data = rhc)

    outcome_full <- lm(make_lm_formula(outcome_var, c(treat_var, covariates)), data = rhc)
    outcome_reduced <- lm(make_lm_formula(outcome_var, c(treat_var, other_covariates)), data = rhc)

    data.table(
        benchmark_covariate = var,
        partial_r2_with_treatment = partial_r2_from_models(treatment_full, treatment_reduced),
        partial_r2_with_outcome = partial_r2_from_models(outcome_full, outcome_reduced)
    )
}))

benchmark_r2[, treatment_outcome_partial_r2 := treatment_outcome_partial_r2] |> invisible()
benchmark_r2[order(-partial_r2_with_treatment, -partial_r2_with_outcome)]
```

```{r partial-r2-plot, fig.width=7, fig.height=5}
benchmark_r2_long <- melt(
    benchmark_r2,
    id.vars = "benchmark_covariate",
    measure.vars = c("partial_r2_with_treatment", "partial_r2_with_outcome"),
    variable.name = "relationship",
    value.name = "partial_r2"
)
benchmark_r2_long[, benchmark_covariate := factor(
    benchmark_covariate,
    levels = benchmark_r2[order(partial_r2_with_treatment), benchmark_covariate]
)] |> invisible()

partial_r2_plot <- ggplot(
    benchmark_r2_long,
    aes(x = benchmark_covariate, y = partial_r2, color = relationship)
) +
    geom_point(size = 2.2) +
    geom_hline(
        yintercept = treatment_outcome_partial_r2,
        color = "grey45",
        linetype = "dashed"
    ) +
    coord_flip() +
    labs(
        x = NULL,
        y = "Partial R-squared",
        color = "Relationship",
        title = "Observed-covariate partial R-squared benchmarks"
    ) +
    theme_minimal()

partial_r2_plot
```

## Findings

- PS trimming does not affect the results much, the ATE and ATT estimation are stable.
- Adjustment sets seems to be best for basic adjustment set.
- E-values summarize how strong an unmeasured confounder would need to be, on the risk-ratio scale, to explain away each estimate.
- Partial R-squared values benchmark unmeasured confounding against observed covariates; larger values mean a covariate explains more residual variation in treatment or outcome.
